Rare-event sampling applied to the simulation of extreme mechanical efforts exerted by a turbulent flow on a bluff body

### 2020-01-07 jfm/draft_p2_jfm.tex
Add jfm styling

\documentclass[pre,aps,floatfix,10pt,superscriptaddress, notitlepage,preprint]{revtex4-1}
\pdfoutput=1
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage[svgnames]{xcolor}
\usepackage{graphicx}
\usepackage{acro}
\usepackage{subfig}
\usepackage{caption}
\graphicspath{{figures/}}

@@

\definecolor{myred}{HTML}{b30000}
\definecolor{mygreen}{HTML}{008000}
\newcommand{\EL}[1]{{\color{myred}{#1}}}
\newcommand{\TL}[1]{{\color{mygreen}{#1}}}

\begin{document}

@@

\author{Thibault Lestang}
\email{thibault.lestang@ens-lyon.fr}
\affiliation{Univ Lyon, ENS de Lyon, Univ Claude Bernard de Lyon, CNRS, Laboratoire de Physique, F-69342 Lyon, France}
\affiliation{Univ Lyon, Ecole Centrale de Lyon, Univ Claude Bernard de Lyon, INSA de Lyon, CNRS, Laboratoire de M\'ecanique des Fluides et d'Acoustique, F-69134 Ecully cedex, France}
\author{Freddy Bouchet}
\email{freddy.bouchet@ens-lyon.fr}
\affiliation{Univ Lyon, ENS de Lyon, Univ Claude Bernard de Lyon, CNRS, Laboratoire de Physique, F-69342 Lyon, France}
\author{Emmanuel Lévêque}
\email{emmanuel.leveque@ec-lyon.fr}
\affiliation{Univ Lyon, Ecole Centrale de Lyon, Univ Claude Bernard de Lyon, INSA de Lyon, CNRS, Laboratoire de M\'ecanique des Fluides et d'Acoustique, F-69134 Ecully cedex, France}

@@

% general comments on physical problem %
%
Turbulent flows are important in a variety of natural phenomena, industrial and civil applications.
Their characteristic feature is the spontaneous development of intense and sporadic motions associated with extreme internal forces \cite{lesieur_book,donzis_sreenivasan_2010,Yeung}.
``Extreme'' refers here to fluctuations that can deviate from the mean value by ${\cal{O}}(10)$ standard deviations.
In engineering, the nature of such extreme dynamical events and their statistics are of crucial interest to predict excessive mechanical efforts.
Such anomalous constraints can threaten the structural integrity of embedded structures.
%

From the viewpoint of chaotic dynamical systems, turbulence in fluids is linked to non-linearity and strong departure from statistical equilibrium \cite{KRAICHNAN}.
The use of perturbative methods in identifying resonant interactions among degrees of freedom responsible for extreme fluctuations is unsuccessful.
Therefore, simulation offers a practical approach to gain physical insight into these events, quantifying their intensity and estimating their frequency of occurrence.
However, this requires very long simulations since these events are rare.
% general def on rare-event sampling %
%
{Rare-event sampling refer to a large body of methods that aim at exploring preferentially regions of phase space corresponding to rare events, that would otherwise be accessed with a very low probability through a brute-force direct sampling.}
%
In the present work, a computational study of extreme mechanical efforts acting on an immersed bluff body is conducting by using both very long time-series (direct sampling) and rare-event sampling techniques.
%

% next two paragraphs may be too technical -- Find a way to summarize them in a few sentences accessible to a broad audience in fluid mechanics? Notion of action will be vague for many readers
%
In fluid turbulence, rare-event sampling has been approached mainly from the perspective of simplified dynamics such as the one-dimensional Burgers' equation with a stochastic forcing \cite{bec_burgers_2007}. In this case, dynamics can be sampled by using a Markov chain Monte-Carlo algorithm \cite{duben_monte_2008,mesterhazy2011anomalous,mesterhazy2013lattice} that provides a framework for rare-event sampling.
%
An alternative approach is based on instantons \cite{gurarie_instantons_1996,grafke2015instanton} and applies to stochastically driven systems in the limit of weak noise.
Instantons refer to the most probable trajectories in phase space that achieve a given rare event (in the limit of weak noise). Suitable numerical schemes can be used to evaluate instantons as well as the related probabilities of rare events \cite{chernykh_large_2001,grafke_instanton_2013,grigorio_instantons_2017,laurie2015computation,bouchet2014langevin}.
An example is the investigation of the physics of rogue waves~\cite{dematteis2018rogue,dematteis2019experimental}.
%
A drawback of the aforementioned approaches is their limitation to simple and stochastically driven dynamics.
%

%
In this paper, a more general approach is considered for complex, possibly deterministic, dynamical systems.
It is based on sampling algorithms relying on \emph{selection rules} applied to an ensemble of trajectories, and is designed to sample rare events of some observable with a higher frequency.
%    
%
Even though such ideas date back to the early 1950s, they have received ever-growing interest over the the last twenty years with successful applications in various domains such as chemistry \cite{van_erp_elaborating_2005,escobedo_transition_2009,teo_adaptive_2016}, biophysics \cite{huber_weighted-ensemble_1996,zuckerman2017weighted,bolhuis2005kinetic}, nuclear physics \cite{louvin2017}, nonlinear dynamical systems \cite{tailleur_probing_2007} and communication networks simulation \cite{villen-altamirano_restart:_1994}.
More importantly, these type of algorithms have been shown to be useful for the study of rare events in simple deterministic dynamics~\cite{wouters2016rare}.
%
%
An original contribution of the present work is certainly the application of rare-event sampling algorithms in the context of far-from-equilibrium dynamics with an irreducible very large number of degrees of freedom. 
%Let us remark that most of previous successful applications involve dynamical systems with low-dimensional attractors. \EL{is it TRUE? \#11}
% 
Two different algorithms suitable for out-of-equilibrium dynamics are considered and compared. Namely, the Adaptive Multilevel Splitting algorithm and the Giardina-Kurchan-Tailleur-Lecomte algorithm.
%

The \acl{ams} algorithm \cite{cerou_adaptive_2007} builds on previous ideas about splitting approaches \cite{KahnHarris1951,glasserman_look_1998,glasserman_multilevel_1999} -- a detailed description will be given later.  
In recent years, it has allowed for the computation of rare events in problems involving a large number of degrees of freedom such as molecular dynamics simulations \cite{aristoff_adaptive_2015,teo_adaptive_2016}. The first use of the \ac{ams} algorithms for more complex dynamics, for instance stochastic partial differential equations was for the computation of rare trajectories in the Allen-Cahn equations \cite{rolland_computing_2016}. More recently is has been applied to rare events in stochastic models of wall-turbulence \cite{rolland_extremely_2018} and atmospheric dynamics \cite{bouchet2019rare}.

During the last decade, the main theoretical framework for the study of rare events in statistical physics has been the theory of large deviations \cite{touchette_large_2009}.
Alongside numerical methods have been developed to sample rare events \cite{moral_feynman-kac_2004}. 
Among them the \ac{gktl} algorithm \cite{giardina_direct_2006} is particularly suited for (chaotic) dynamical systems \cite{giardina_simulating_2011,Laffargue_2013}.
%
{Recently the \ac{gktl} algorithm has allowed to successfully perform the numerical simulation of extreme heat waves in a simplified modelling of the atmosphere \cite{ragone_computation_2018}. 
This achievement represents a significant leap in the applicability of rare-event sampling to complex dynamical systems. Along the same line, rare-event sampling algorithms are here pushed aside traditional applications to consider fluid-structure interaction in a turbulent flow.}
%

@@

% geometry
%
The simulated flow develops in a long plane channel of dimension $513 \times 129$ mesh points. The square obstacle has size $R=16$ (in mesh unit) and is located at the centre of the channel. The spacing and bar height of the entrance grid are both equal to $R/2$ (see Fig.~\ref{fig:illustr_ecoulement}). 
% boundary conditions
%
No-slip boundary conditions are enforced on top and bottom walls of the channel and on the surface of the obstacle by using an halfway bounce-back procedure~\cite{lbm_book}.  
%
Upstream of the grid, a constant parabolic velocity profile and a constant mass density (equal to unity) are imposed as an inlet condition. 
The centerline velocity is $0.05$ in lattice units, \textit{i.e.} normalised by $\Delta x$ and $\Delta t$ referring to the lattice resolution and the time-step respectively. The initial distributions are imposed at equilibrium (see Appendix). 
In the bulk, the viscosity is adjusted so that grid turbulence is generated with Reynolds number $\mathrm{Re_{grid}}=1200$. The reference Mach number is equal to $0.06$ in agreement with the assumption of weak compressibility of the \ac{lbm}. 
Near the end of the channel, the flow is progressively damped within a \textit{sponge layer} where the viscosity is artificially enhanced.
Finally, the outlet boundary condition relies on a second-order extrapolation of the velocity and mass density.
The extrapolated distributions are evaluated through a regularization procedure relying on a finite difference estimation of the local stress tensor, as introduced in \cite{latt2008straight}.

@@

The waiting times $\tau$ are defined as the time between two consecutive occurrences of peak fluctuations with amplitude $f_d \geq a$, as illustrated in Fig.~\ref{fig:illustrate_return_time}.
The mixing time $\tau_m$ is the time needed for the dynamics to loose the memory of its initial conditions.
As soon as the typical waiting times are much larger than the mixing time $\tau_m$, the occurrences of such events follow a Poisson process and the distribution of the waiting times is exponential, \emph{i.e.} $P(\tau)=\lambda(a)\exp(-\lambda(a)\tau)$ where $r(a)=1/\lambda(a)$ is the averaged waiting time \cite{lestang_computing_2018}; $r(a)$ is called the {\it return time} of the level $a$.
For systems without multi-stability, it is common for the mixing time $\tau_m$ to be of the order of the correlation time $\tau_c$.

How rare is a fluctuation $a$ is quantified by its return time $r(a)$.
We can define extreme drag fluctuations as \textit{rare events} in the sense that the return time is much larger than the correlation time, \emph{i.e.} $r(a) \gg \tau_c$.
%
%
%The shape of the return time $r$ as a function of the return level $a$ is system dependent. 
If one assumes that  $r(a) = t(a)~/~\mathbb{P}(f_d\geq a)$   where the time scale $t(a)$ is of order $\tau_c$ and varies much more slowly with $a$ than ${\mathbb{P}(f_d\geq a)}$,
% As the PDF ${\mathbb{P}(f_d\geq a)}$ was shown to be well approximated by an exponential (see figure Fig.~\ref{fig:pdf_drag})
one might expect that 
\begin{equation}
 \label{eq:return_time}
r(a) \underset{a\to\infty}{\propto} \exp(-la)
\end{equation}
where $l$ is the rate describing the positive tail of the \ac{pdf} of the drag (shown in Fig.~\ref{fig:pdf_drag}).
Fig.~\ref{fig:return_time_instant} shows the evolution of the return time $r(a)$ with the amplitude of fluctuation $a$, computed from {direct sampling} of the drag signal $f_d(t)$~\cite{lestang_computing_2018}.
Consistently, it is found that the return time $r(a)$ is well approximated by an exponential for large levels $a$. Let us also point out some deviation from the exponential law at the largest levels, which are probably the consequence of under-sampling.

@@

% neither case (1) nor case (2)
%
Fig.~\ref{fig:extreme_avg} displays the time-series $\{f_d(t)\}_{t^{\star} \leq t \leq t^{\star}+T}$ for several extreme fluctuations of $F_T$ occurring at $t=t^\star$.
It is found that extreme fluctuations of the time-averaged drag can neither be reduced to case (1) nor case (2). 
Indeed, both cases are equally featured in Fig.~\ref{fig:extreme_avg}.
Very large value of the averaged-drag results from either a very large fluctuation, or a significant succession of moderate (positive) fluctuations of the instantaneous drag.
%
% theoretical explanation
%
This observation can be related to the exponential shape of the tail of the \ac{pdf} describing extreme positive drag fluctuations. 
Indeed, let $X$ be a random variable with a \ac{pdf} $\mathbb{P}(X)$ and a standard deviation $\sigma_X$.
Considering an extreme positive value of $S_N=\sum_{n=1}^{N}X_n$, the probability $p_1$ (resp. $p_2$) of case (1) (resp. case (2)) writes   
\begin{equation}
  \label{eq:indep}
  p_{1}  \left(\sum_{n=1}^{N} X_n = S_N\right) \propto \mathbb{P} \left({S_N}/{N} \right)^{N} \quad \text{and} \quad p_{2}  \left(\sum_{n=1}^{N} X_n= S_N\right) \propto \mathbb{P}(S_N)  
\end{equation}
If $\mathbb{P}$ has an exponential positive tail, \textit{i.e.} $\mathbb{P}(X=x) \underset{x \gg \sigma_X}{\propto} e^{-\lambda x}$, both cases (1) and (2) have equivalent probabilities provided that the average $a=S_N/N$ is very large:
\begin{equation}
  \frac{p_{2}}{p_{1}} \underset{a\to \infty}{\to} \left(e^{\lambda a }\right)^{N} e^{-\lambda a N} = 1
  \label{eq:ratio_exp}
\end{equation}
The observation is therefore well supported theoretically.
See~\cite{lestang2019JSTAT} for further details.

@@

% Introduction of AMS
%
The \acl{ams} algorithm~\cite{cerou_adaptive_2007} builds on previous ideas of {splitting algorithms} \cite{KahnHarris1951}.
The sampling of a rare event is made easier by splitting the dynamical path of the system (in phase space) into 
%a sequence of events that can be sampled with a higher probability  
a sequence of shorter excursions \cite{glasserman_multilevel_1999,rolland_statistical_2015}.
The related selection-mutation step of this algorithm {relies on the definition of a \textit{score function} $\xi (\mathbf{x}(t),t)$}.
The operating principle is sketched in Fig.~\ref{fig:illustr_AMS}.
%
% As already mentioned in the introduction, the \ac{ams} has been successfully applied to a large body of problems for the past ten years. 
%
% Over the past ten years the \ac{ams} has been successfully applied to a large body of problems, including the computation of the dissociation time of biomolecules~\cite{teo_adaptive_2016} or the simulation of rare relaminarizations in a stochastic model of wall turbulence~\cite{rolland_extremely_2018}.
%
% Quick description of TAMS
%
Iterations of the algorithm consist in discarding the trajectories with the lowest maxima of the score function during the time interval $0\leq t \leq T_a$.
Discarded trajectories are re-sampled according to the remaining trajectories.
See~\cite{cerou2019adaptive} for a review of the \ac{ams} algorithm, its history and applications.
For recent applications of the \ac{ams} in the context of fluid turbulence, see \cite{bouchet2019rare,rolland_extremely_2018}.
{Let us mention that we use a specific form of the \ac{ams}, in which trajectories have a fixed duration $T_a$.
  It can however be formulated in terms of the ``classical \ac{ams}''~\cite{lestang_computing_2018}.
  The following results are not limited to this particular variant.}

\subsubsection{Illustration of the \ac{ams} on a simple case: the \acl{ou} process}
\label{sec:ams_ou}
% Example: OU process
\begin{figure}
  \centering
  \includegraphics[width=.7\linewidth]{AMS_OU/AMS_OU.png}
    \caption{Efficiency of the \ac{tams} algorithm with respect to direct sampling in the case of an Ornstein-Uhlenbeck process \cite{lestang_computing_2018}. The red line represents the evolution of the maximum obtained from re-sampled trajectories as a function of the computational cost $C_{TAMS}$. The blue line is the analytical solution for the return time of amplitude $a$.}
  \label{fig:comparaison_temps_de_retour}
\end{figure}
Fluid dynamics is temporarily left aside and a one-dimensional \acl{ou} process is considered:
\begin{equation}
  \label{eq:ou}
  \dot{x} = -x + \eta (t),
\end{equation}
where $\eta$ is a Gaussian noise with $\langle \eta(t)\eta(t-t')\rangle = \delta(t-t')$.
This basic stochastic process will allow us to highlight differences with fluid dynamics.

@@

% Application to instantaneous drag
%
% Can the \ac{ams} algorithm achieve similar performance when applied to complex chaotic dynamics, such as turbulent flows around objects ?
% To answer this question, we consider again the two-dimensional test flow introduced in section~\ref{sec:test_flow}.
%
Our aim is now to use the \ac{ams} algorithm to sample flow evolutions that exhibit extreme fluctuations of the drag $f_d$ acting on the square obstacle.
In contrast with the \acl{ou} process, the phase space is here highly-dimensional with intrinsically chaotic dynamics.
In that situation, the choice of the score function $\xi (\mathbf{x}(t),t)$ is no longer straightforward.
However a natural simple choice is $\xi (\mathbf{x}(t),t) = f_d(t)$, for which we opt for in the following.
%
{Let us also stress that the dynamics is here deterministic.
Therefore, randomness must be artificially introduced for the re-sampling procedure to be effective, \textit{i.e.} to generate new trajectories associated to larger drag fluctuations.
In practice, a small perturbation in the state of the flow has been introduced at the branching points (see Fig.~\ref{fig:illustr_AMS} and~\cite{wouters2016rare}).
%
Due to the chaotic behavior of the flow, re-sampled trajectories then separate from their parents.
% The timescale over which two trajectories separate due to an infinitesimal perturbation is known as the \textit{Lyapunov timescale}. It is denoted $\tau_L$ in the remainder of this paper.
We assume that this perturbation at branching times has a negligible impact on the statistical properties of the sampled rare events.
The validity of this assumption has been tested by performing a long simulation of the dynamics and by regularly perturbing the flow.
It was checked that the obtained statistics of the drag were consistent with the statistics computed from the (unperturbed) control simulation described in section~\ref{sec:pdfs}.}
%\EL{Freddy: provide reference for Lyapunov timescale. \#12}

@@

In the following, the \ac{gktl} algorithm is used with $N=1024$ trajectories of duration $T_a > T= 10\tau_c$.
Each trajectory in the biased ensemble results in a timeseries of the time-averaged drag
\begin{equation}
\label{eq:time_averaged}
F_T(t) = \int_{t-T}^{t}f_d(\tau)d\tau, \quad t\in [T,T_a]  
\end{equation}
and the return time of a fluctuation $F_T \geq a$ is given by~\cite{lestang_computing_2018}
\begin{equation}
r(a) = - \frac{T_a - T}{\ln (1-\mathbb{P}(F_T \geq a))}.
\end{equation}

>>>>

\documentclass{jfm}
\pdfoutput=1
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage[svgnames]{xcolor}
\usepackage{graphicx}
\usepackage{acro}
\usepackage{subfig}
\usepackage{caption}
\graphicspath{{../figures/}}

@@

@@

\author{Thibault Lestang\aff{1}\aff{2}
  \corresp{\email{thibault.lestang@cs.ox.ac.uk}},
  Freddy Bouchet\aff{1}
  \and Emmanuel L\'evêque\aff{2}}
\affiliation{\aff{1}Univ Lyon, ENS de Lyon, Univ Claude Bernard de Lyon, CNRS, Laboratoire de Physique, F-69342 Lyon, France
\aff{2}Univ Lyon, Ecole Centrale de Lyon, Univ Claude Bernard de Lyon, INSA de Lyon, CNRS, Laboratoire de M\'ecanique des Fluides et d'Acoustique, F-69134 Ecully cedex, France}
\begin{document}

@@

% general comments on physical problem %

Turbulent flows are important in a variety of natural phenomena, industrial and civil applications.
Their characteristic feature is the spontaneous development of intense and sporadic motions associated with extreme internal forces \citep{lesieur_book,donzis_sreenivasan_2010,Yeung}.
``Extreme'' refers here to fluctuations that can deviate from the mean value by ${\cal{O}}(10)$ standard deviations.
In engineering, the nature of such extreme dynamical events and their statistics are of crucial interest to predict excessive mechanical efforts.
Such anomalous constraints can threaten the structural integrity of embedded structures.
%

From the viewpoint of chaotic dynamical systems, turbulence in fluids is linked to non-linearity and strong departure from statistical equilibrium \citep{KRAICHNAN}.
The use of perturbative methods in identifying resonant interactions among degrees of freedom responsible for extreme fluctuations is unsuccessful.
Therefore, simulation offers a practical approach to gain physical insight into these events, quantifying their intensity and estimating their frequency of occurrence.
However, this requires very long simulations since these events are rare.
% general def on rare-event sampling %
%
{Rare-event sampling refer to a large body of methods that aim at exploring preferentially regions of phase space corresponding to rare events, that would otherwise be accessed with a very low probability through a brute-force direct sampling.}
%
In the present work, a computational study of extreme mechanical efforts acting on an immersed bluff body is conducting by using both very long time-series (direct sampling) and rare-event sampling techniques.
%

% next two paragraphs may be too technical -- Find a way to summarize them in a few sentences accessible to a broad audience in fluid mechanics? Notion of action will be vague for many readers
%
In fluid turbulence, rare-event sampling has been approached mainly from the perspective of simplified dynamics such as the one-dimensional Burgers' equation with a stochastic forcing \citep{bec_burgers_2007}. In this case, dynamics can be sampled by using a Markov chain Monte-Carlo algorithm \citep{duben_monte_2008,mesterhazy2011anomalous,mesterhazy2013lattice} that provides a framework for rare-event sampling.
%
An alternative approach is based on instantons \citep{gurarie_instantons_1996,grafke2015instanton} and applies to stochastically driven systems in the limit of weak noise.
Instantons refer to the most probable trajectories in phase space that achieve a given rare event (in the limit of weak noise). Suitable numerical schemes can be used to evaluate instantons as well as the related probabilities of rare events \citep{chernykh_large_2001,grafke_instanton_2013,grigorio_instantons_2017,laurie2015computation,bouchet2014langevin}.
An example is the investigation of the physics of rogue waves~\citep{dematteis2018rogue,dematteis2019experimental}.
%
A drawback of the aforementioned approaches is their limitation to simple and stochastically driven dynamics.
%

%
In this paper, a more general approach is considered for complex, possibly deterministic, dynamical systems.
It is based on sampling algorithms relying on \emph{selection rules} applied to an ensemble of trajectories, and is designed to sample rare events of some observable with a higher frequency.
%    
%
Even though such ideas date back to the early 1950s, they have received ever-growing interest over the the last twenty years with successful applications in various domains such as chemistry \citep{van_erp_elaborating_2005,escobedo_transition_2009,teo_adaptive_2016}, biophysics \citep{huber_weighted-ensemble_1996,zuckerman2017weighted,bolhuis2005kinetic}, nuclear physics \citep{louvin2017}, nonlinear dynamical systems \citep{tailleur_probing_2007} and communication networks simulation \citep{villen-altamirano_restart:_1994}.
More importantly, these type of algorithms have been shown to be useful for the study of rare events in simple deterministic dynamics~\citep{wouters2016rare}.
%
%
An original contribution of the present work is certainly the application of rare-event sampling algorithms in the context of far-from-equilibrium dynamics with an irreducible very large number of degrees of freedom. 
%Let us remark that most of previous successful applications involve dynamical systems with low-dimensional attractors. \EL{is it TRUE? \#11}
% 
Two different algorithms suitable for out-of-equilibrium dynamics are considered and compared. Namely, the Adaptive Multilevel Splitting algorithm and the Giardina-Kurchan-Tailleur-Lecomte algorithm.
%

The \acl{ams} algorithm \citep{cerou_adaptive_2007} builds on previous ideas about splitting approaches \citep{KahnHarris1951,glasserman_look_1998,glasserman_multilevel_1999} -- a detailed description will be given later.  
In recent years, it has allowed for the computation of rare events in problems involving a large number of degrees of freedom such as molecular dynamics simulations \citep{aristoff_adaptive_2015,teo_adaptive_2016}. The first use of the \ac{ams} algorithms for more complex dynamics, for instance stochastic partial differential equations was for the computation of rare trajectories in the Allen-Cahn equations \citep{rolland_computing_2016}. More recently is has been applied to rare events in stochastic models of wall-turbulence \citep{rolland_extremely_2018} and atmospheric dynamics \citep{bouchet2019rare}.

During the last decade, the main theoretical framework for the study of rare events in statistical physics has been the theory of large deviations \citep{touchette_large_2009}.
Alongside numerical methods have been developed to sample rare events \citep{moral_feynman-kac_2004}. 
Among them the \ac{gktl} algorithm \citep{giardina_direct_2006} is particularly suited for (chaotic) dynamical systems \citep{giardina_simulating_2011,Laffargue_2013}.
%
{Recently the \ac{gktl} algorithm has allowed to successfully perform the numerical simulation of extreme heat waves in a simplified modelling of the atmosphere \citep{ragone_computation_2018}. 
This achievement represents a significant leap in the applicability of rare-event sampling to complex dynamical systems. Along the same line, rare-event sampling algorithms are here pushed aside traditional applications to consider fluid-structure interaction in a turbulent flow.}
%

@@

% geometry
%
The simulated flow develops in a long plane channel of dimension $513 \times 129$ mesh points. The square obstacle has size $R=16$ (in mesh unit) and is located at the centre of the channel. The spacing and bar height of the entrance grid are both equal to $R/2$ (see Fig.~\ref{fig:illustr_ecoulement}). 
% boundary conditions
%
No-slip boundary conditions are enforced on top and bottom walls of the channel and on the surface of the obstacle by using an halfway bounce-back procedure~\citep{lbm_book}.  
%
Upstream of the grid, a constant parabolic velocity profile and a constant mass density (equal to unity) are imposed as an inlet condition. 
The centerline velocity is $0.05$ in lattice units, \textit{i.e.} normalised by $\Delta x$ and $\Delta t$ referring to the lattice resolution and the time-step respectively. The initial distributions are imposed at equilibrium (see Appendix). 
In the bulk, the viscosity is adjusted so that grid turbulence is generated with Reynolds number $\mathrm{Re_{grid}}=1200$. The reference Mach number is equal to $0.06$ in agreement with the assumption of weak compressibility of the \ac{lbm}. 
Near the end of the channel, the flow is progressively damped within a \textit{sponge layer} where the viscosity is artificially enhanced.
Finally, the outlet boundary condition relies on a second-order extrapolation of the velocity and mass density.
The extrapolated distributions are evaluated through a regularization procedure relying on a finite difference estimation of the local stress tensor, as introduced in \citep{latt2008straight}.

@@

The waiting times $\tau$ are defined as the time between two consecutive occurrences of peak fluctuations with amplitude $f_d \geq a$, as illustrated in Fig.~\ref{fig:illustrate_return_time}.
The mixing time $\tau_m$ is the time needed for the dynamics to loose the memory of its initial conditions.
As soon as the typical waiting times are much larger than the mixing time $\tau_m$, the occurrences of such events follow a Poisson process and the distribution of the waiting times is exponential, \emph{i.e.} $P(\tau)=\lambda(a)\exp(-\lambda(a)\tau)$ where $r(a)=1/\lambda(a)$ is the averaged waiting time \citep{lestang_computing_2018}; $r(a)$ is called the {\it return time} of the level $a$.
For systems without multi-stability, it is common for the mixing time $\tau_m$ to be of the order of the correlation time $\tau_c$.

How rare is a fluctuation $a$ is quantified by its return time $r(a)$.
We can define extreme drag fluctuations as \textit{rare events} in the sense that the return time is much larger than the correlation time, \emph{i.e.} $r(a) \gg \tau_c$.
%
%
%The shape of the return time $r$ as a function of the return level $a$ is system dependent. 
If one assumes that  $r(a) = t(a)~/~\mathbb{P}(f_d\geq a)$   where the time scale $t(a)$ is of order $\tau_c$ and varies much more slowly with $a$ than ${\mathbb{P}(f_d\geq a)}$,
% As the PDF ${\mathbb{P}(f_d\geq a)}$ was shown to be well approximated by an exponential (see figure Fig.~\ref{fig:pdf_drag})
one might expect that 
\begin{equation}
 \label{eq:return_time}
r(a) \underset{a\to\infty}{\propto} \exp(-la)
\end{equation}
where $l$ is the rate describing the positive tail of the \ac{pdf} of the drag (shown in Fig.~\ref{fig:pdf_drag}).
Fig.~\ref{fig:return_time_instant} shows the evolution of the return time $r(a)$ with the amplitude of fluctuation $a$, computed from {direct sampling} of the drag signal $f_d(t)$~\citep{lestang_computing_2018}.
Consistently, it is found that the return time $r(a)$ is well approximated by an exponential for large levels $a$. Let us also point out some deviation from the exponential law at the largest levels, which are probably the consequence of under-sampling.

@@

% neither case (1) nor case (2)
%
Fig.~\ref{fig:extreme_avg} displays the time-series $\{f_d(t)\}_{t^{\star} \leq t \leq t^{\star}+T}$ for several extreme fluctuations of $F_T$ occurring at $t=t^\star$.
It is found that extreme fluctuations of the time-averaged drag can neither be reduced to case (1) nor case (2). 
Indeed, both cases are equally featured in Fig.~\ref{fig:extreme_avg}.
Very large value of the averaged-drag results from either a very large fluctuation, or a significant succession of moderate (positive) fluctuations of the instantaneous drag.
%
% theoretical explanation
%
This observation can be related to the exponential shape of the tail of the \ac{pdf} describing extreme positive drag fluctuations. 
Indeed, let $X$ be a random variable with a \ac{pdf} $\mathbb{P}(X)$ and a standard deviation $\sigma_X$.
Considering an extreme positive value of $S_N=\sum_{n=1}^{N}X_n$, the probability $p_1$ (resp. $p_2$) of case (1) (resp. case (2)) writes   
\begin{equation}
  \label{eq:indep}
  p_{1}  \left(\sum_{n=1}^{N} X_n = S_N\right) \propto \mathbb{P} \left({S_N}/{N} \right)^{N} \quad \text{and} \quad p_{2}  \left(\sum_{n=1}^{N} X_n= S_N\right) \propto \mathbb{P}(S_N)  
\end{equation}
If $\mathbb{P}$ has an exponential positive tail, \textit{i.e.} $\mathbb{P}(X=x) \underset{x \gg \sigma_X}{\propto} e^{-\lambda x}$, both cases (1) and (2) have equivalent probabilities provided that the average $a=S_N/N$ is very large:
\begin{equation}
  \frac{p_{2}}{p_{1}} \underset{a\to \infty}{\to} \left(e^{\lambda a }\right)^{N} e^{-\lambda a N} = 1
  \label{eq:ratio_exp}
\end{equation}
The observation is therefore well supported theoretically.
See~\citep{lestang2019JSTAT} for further details.

@@

% Introduction of AMS
%
The \acl{ams} algorithm~\citep{cerou_adaptive_2007} builds on previous ideas of {splitting algorithms} \citep{KahnHarris1951}.
The sampling of a rare event is made easier by splitting the dynamical path of the system (in phase space) into 
%a sequence of events that can be sampled with a higher probability  
a sequence of shorter excursions \citep{glasserman_multilevel_1999,rolland_statistical_2015}.
The related selection-mutation step of this algorithm {relies on the definition of a \textit{score function} $\xi (\mathbf{x}(t),t)$}.
The operating principle is sketched in Fig.~\ref{fig:illustr_AMS}.
%
% As already mentioned in the introduction, the \ac{ams} has been successfully applied to a large body of problems for the past ten years. 
%
% Over the past ten years the \ac{ams} has been successfully applied to a large body of problems, including the computation of the dissociation time of biomolecules~\citep{teo_adaptive_2016} or the simulation of rare relaminarizations in a stochastic model of wall turbulence~\citep{rolland_extremely_2018}.
%
% Quick description of TAMS
%
Iterations of the algorithm consist in discarding the trajectories with the lowest maxima of the score function during the time interval $0\leq t \leq T_a$.
Discarded trajectories are re-sampled according to the remaining trajectories.
See~\citep{cerou2019adaptive} for a review of the \ac{ams} algorithm, its history and applications.
For recent applications of the \ac{ams} in the context of fluid turbulence, see \citep{bouchet2019rare,rolland_extremely_2018}.
{Let us mention that we use a specific form of the \ac{ams}, in which trajectories have a fixed duration $T_a$.
  It can however be formulated in terms of the ``classical \ac{ams}''~\citep{lestang_computing_2018}.
  The following results are not limited to this particular variant.}

\subsubsection{Illustration of the \ac{ams} on a simple case: the \acl{ou} process}
\label{sec:ams_ou}
% Example: OU process
\begin{figure}
  \centering
  \includegraphics[width=.7\linewidth]{AMS_OU/AMS_OU.png}
    \caption{Efficiency of the \ac{tams} algorithm with respect to direct sampling in the case of an Ornstein-Uhlenbeck process \citep{lestang_computing_2018}. The red line represents the evolution of the maximum obtained from re-sampled trajectories as a function of the computational cost $C_{TAMS}$. The blue line is the analytical solution for the return time of amplitude $a$.}
  \label{fig:comparaison_temps_de_retour}
\end{figure}
Fluid dynamics is temporarily left aside and a one-dimensional \acl{ou} process is considered:
\begin{equation}
  \label{eq:ou}
  \dot{x} = -x + \eta (t),
\end{equation}
where $\eta$ is a Gaussian noise with $\langle \eta(t)\eta(t-t')\rangle = \delta(t-t')$.
This basic stochastic process will allow us to highlight differences with fluid dynamics.

@@

% Application to instantaneous drag
%
% Can the \ac{ams} algorithm achieve similar performance when applied to complex chaotic dynamics, such as turbulent flows around objects ?
% To answer this question, we consider again the two-dimensional test flow introduced in section~\ref{sec:test_flow}.
%
Our aim is now to use the \ac{ams} algorithm to sample flow evolutions that exhibit extreme fluctuations of the drag $f_d$ acting on the square obstacle.
In contrast with the \acl{ou} process, the phase space is here highly-dimensional with intrinsically chaotic dynamics.
In that situation, the choice of the score function $\xi (\mathbf{x}(t),t)$ is no longer straightforward.
However a natural simple choice is $\xi (\mathbf{x}(t),t) = f_d(t)$, for which we opt for in the following.
%
{Let us also stress that the dynamics is here deterministic.
Therefore, randomness must be artificially introduced for the re-sampling procedure to be effective, \textit{i.e.} to generate new trajectories associated to larger drag fluctuations.
In practice, a small perturbation in the state of the flow has been introduced at the branching points (see Fig.~\ref{fig:illustr_AMS} and~\citep{wouters2016rare}).
%
Due to the chaotic behavior of the flow, re-sampled trajectories then separate from their parents.
% The timescale over which two trajectories separate due to an infinitesimal perturbation is known as the \textit{Lyapunov timescale}. It is denoted $\tau_L$ in the remainder of this paper.
We assume that this perturbation at branching times has a negligible impact on the statistical properties of the sampled rare events.
The validity of this assumption has been tested by performing a long simulation of the dynamics and by regularly perturbing the flow.
It was checked that the obtained statistics of the drag were consistent with the statistics computed from the (unperturbed) control simulation described in section~\ref{sec:pdfs}.}
%\EL{Freddy: provide reference for Lyapunov timescale. \#12}

@@

In the following, the \ac{gktl} algorithm is used with $N=1024$ trajectories of duration $T_a > T= 10\tau_c$.
Each trajectory in the biased ensemble results in a timeseries of the time-averaged drag
\begin{equation}
\label{eq:time_averaged}
F_T(t) = \int_{t-T}^{t}f_d(\tau)d\tau, \quad t\in [T,T_a]  
\end{equation}
and the return time of a fluctuation $F_T \geq a$ is given by~\citep{lestang_computing_2018}
\begin{equation}
r(a) = - \frac{T_a - T}{\ln (1-\mathbb{P}(F_T \geq a))}.
\end{equation}
